ExpressionView - an interactive viewer for biclusters identified in gene expression data

### 2009-09-29 expressionviewnote/expressionview.tex

To determine the efficiency of this method, we have studied a large number of perfectly orderable, but initially scrambled, test cases. We find that in almost every situation, the proposed algorithm finds an order that recovers more than 99\% of the score of the optimal solution and in most cases, it actually recovers the correct alignment. In these test samples, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [4, 4.5]$, almost independent of the number of elements $n$.

@@

\subsection{Visualisation}
The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data and the biclusters. Relying on the Flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights. A screenshot is shown in Fig.~\ref{fig:workflow}. The interface is divided in two parts: On the left-hand side, the user finds the gene expression data in the common heat map form, on top of which the biclusters are overlaid. On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations (GO and KEGG) are shown. Wherever possible, these elements are linked to the corresponding databases. The interface essentially behaves as an image viewer, allowing the user to zoom and pan around the expression data, getting instant feedback on the selected item. The currently selected region can be exported to a PDF file, helpful to prepare figures for presentations or articles, or a simple text file.\\

@@

%\bibliography{../../expressionview}
%\bibliographystyle{plainnat}

@@

@@

\bibitem[Cs{\'a}rdi et~al.(2009)]{csardi09}
Cs{\'a}rdi,G., Kutalik,Z. and Bergmann,S. (2009)
Eisa.
\href{}{\emph{Bioinformatics}, ...}

@@

\bibitem[Kaiser and Leisch(2008)]{kaiser08}
Kaiser,S. and Leisch,F. (2008)
A toolbox for bicluster analysis in r.
\href{http://epub.ub.uni-muenchen.de/3293/}{Technical Report 028, Department of Statistics, University of Munich}.

>>>>

To determine the efficiency of this method, we have studied a large number of perfectly orderable, but initially scrambled, test cases. We find that in almost every situation, the proposed algorithm finds an order that recovers more than 99\% of the score of the optimal solution and in most cases, it actually recovers the correct alignment. For random samples, which are more representative for actual gene expression data than orderable situations, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [1.6, 2]$, almost independent of the number of elements $n$. For a given number of clusters, we find ${\mathcal O}(n^\alpha)$, with $\alpha \in [2.5, 2.7]$.

@@

\subsection{Visualisation}
The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data and the biclusters. Relying on the Flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights. A screenshot is shown in Fig.~\ref{fig:workflow}. The interface is divided in two parts: On the left-hand side, the user finds the gene expression data in the common heat map form, on top of which the biclusters are overlaid. On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations [GO (Gene Ontology)~\citep{ashburner00} and KEGG (Kyoto Encyclopedia of Genes and Genomes)~\citep{kanehisa04}] are shown. Wherever possible, these elements are linked to the corresponding databases. The interface essentially behaves as an image viewer, allowing the user to zoom and pan around the expression data, getting instant feedback on the selected item. The currently selected region can be exported to a PDF file, helpful to prepare figures for presentations or articles, or a simple text file.\\

@@

@@

\bibitem[Ashburner et~al.(2000)]{ashburner00}
Ashburner,M. \emph{et al.} (2000)
Gene Ontology: tool for the unification of biology.
\href{http://dx.doi.org/10.1038/75556}{\emph{Nature Genetics}, {\bf 25}, 25 -- 29.}

@@

\bibitem[Cs{\'a}rdi et~al.(2009)]{csardi09}
Cs{\'a}rdi,G., Kutalik,Z. and Bergmann,S. (2009)
Eisa.
\href{http://dx.doi.org/}{\emph{Bioinformatics}, ...}

@@

\bibitem[Kaiser and Leisch(2008)]{kaiser08}
Kaiser,S. and Leisch,F. (2008)
A toolbox for bicluster analysis in R.
\href{http://epub.ub.uni-muenchen.de/3293/}{Technical Report 028, Department of Statistics, University of Munich}.

\bibitem[Kanehisa et~al.(2004)]{kanehisa04}
Kanehisa,M. \emph{et al.} (2004)
The KEGG resource for deciphering the genome.
\href{http://dx.doi.org/10.1093/nar/gkh063}{\emph{Nucleic Acids Res.}, {\bf 32} (Database special issue), D277.}
